---
title: "Metadata on KarnatakaForest data"
format:
  pdf
---

This is the description of the data KarnatakaForest, that were available in a previous version of the package BIOMASS (v 2.1.11).
You can access these data in a csv file [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/4_tidyverse/data/KarnatakaForest.csv){preview-link="false"}.

# Karnataka forest dataset

## Description

Dataset from 96 forest plots (1 ha) established in the central Western Ghats of India by Ramesh et al. (2010).

## Usage

```{r, eval = FALSE}
data("KarnatakaForest")
```


## Format

A data frame with 65889 observations on the following 8 variables :

* plotId: Names of the plots

* treeId: Tree Id, contains a letter (A, B, C...) when an individual has multiple stems

* family: Family

* genus: Genus

* species: Species

* D: Diameter (cm)

* lat: Latitude

* long: Longitude

## References

Ramesh, B. R. et al. (2010). *Forest stand structure and composition in 96 sites along environmental gradients in the central Western Ghats of India* Ecological Archives E091-216. Ecology, 91(10), 3118-3118.

## Examples

```{r, eval = FALSE}
data(KarnatakaForest)
str(KarnatakaForest)
```

