---
title: "Exercices on session *Getting started with R (2/3)*"
date: 09/04/2025
format:
  pdf
execute:
  echo: false  
---

# RStudio project

* Create a RStudio project to organise the work we will do during the class. Make sure you organise the project in a clear way.

* Create a new script for these exercises (and save it).

# Vectors

Let's work on tree aboveground biomass (AGB). 

We have 2 different forest plots, each containing 4 trees.

* Create two vectors called AGB1 and AGB2 and store the AGB of each of the trees in these plots. Use the following AGB values (in tons).

```{r}
AGB <- data.frame(Plot = c("Plot1", "Plot2"),
                  Tree1 = c(0.1, 0.24),
                  Tree2 = c(0.42, 0.29),
                  Tree3 = c(0.25, 0.2),
                  Tree4 = c(0.3, 0.3))
knitr::kable(AGB)
```

```{r}
AGB1 <- c(0.1, 0.42, 0.25, 0.3)
AGB2 <- c(0.24, 0.29, 0.2, 0.3)
```

* Add names to these vectors (for Tree1, Tree2, etc...)

```{r}
tree_names <- c("Tree1", "Tree2", "Tree3", "Tree4")
names(AGB1) <- tree_names
names(AGB2) <- tree_names
```

* Create two new vectors to store the AGB in kg

```{r}
AGB1_kg <- AGB1 * 1000
AGB2_kg <- AGB2 * 1000
```


* Calculate the mean AGB for each of the plots.

```{r}
mean1 <- mean(AGB1)
mean2 <-mean(AGB2)
```

* Calculate the difference between Tree1 in plot 1 and plot 2, Tree2 in plot 1 and plot 2, etc...

```{r}
delta <- AGB1 - AGB2
```


# Factors

Let's now work on the floristic composition of a plot that has 10 trees of 3 different species.

* Create a vector called *compo* that contains the name of the species of each tree (you can use spA, spB, or the name of you favourite species). 
Your vector should look like this:

```{r}
compo <- c("spA", "spB", "spA", "spC", "spA", "spA", "spC", "spB", "spA", "spC")
compo
```

* Transform *compo* to a factor and check that all is ok (number of trees and number of species).

```{r}
compo <- as.factor(compo)
# str(compo)
# length(compo)
```

* Add a level for a fourth species

```{r}
levels(compo)  <- c(levels(compo), "spD")
```

* Count the number of trees per species

```{r}
tab <- table(compo)
```

