---
title: "Exercices on session *Getting started with R (3/3)*"
format:
  pdf
execute:
  echo: false  
---
  
# Matrices

We are going to use a matrix called *occ* to get data on occurrence of species in sites (*1* if the species is present, and *0* if it is absent).

* Create the following matrix and check that we have the structure we want.

```{r}
# t.seed(22)
# occ <- matrix(sample(c(0,1), 20, replace="TRUE"), 4, 5)
occ <- matrix(c(1, 1, 1, 1, 0, 1, 0, 0, 1, 1, 0, 1, 0, 1, 0, 0, 1, 1, 0, 1), 
              4, 
              5)
colnames(occ) <- c("Sp1", "Sp2", "Sp3", "Sp4", "Sp5")
rownames(occ) <- c("site1", "site2", "site3", "site4")
occ
```
```{r, eval=FALSE}
str(occ)
```
* Calculate the number of species in each site.

```{r, eval=FALSE}
rowSums(occ)
```

* Calculate the number of sites in which each species occurs.

```{r, eval=FALSE}
colSums(occ)
```
* Transpose the matrix to get the following:

```{r}
t(occ)
```

# Lists

* Create the following list:

```{r}
v <- c("site1" = "DF", "site2"= "GO", "site3" = "GO", "site4" = "MT")
dt <- list(occurence = occ,
           site_state = v)
dt
```

# Operators and Functions

* Calculate the square root of 25

```{r, eval=FALSE}
sqrt(25)
```
* Check that the exponential of the logarithm of 1 is equal to 1.
*NB: If you try to do it for another number, you will get an unexpected result due to rounding.*

```{r, eval=FALSE}
exp(log(1)) == 1

# for other values
round(exp(log(3))) == 3
```

* Check that 2 is between 0 and 10 (there are different ways to do it)

```{r, eval=FALSE}
2 %in% 0:10
# or
2 <= 10 & 2 >= 0
```

* Get the absolute value of each element of the following vector *vec*:

```{r}
vec <- c(-4, 5, 8, -9, -3)
vec
```

```{r, eval=FALSE}
abs(vec)
```
* Get the range of values in *vec* 

```{r, eval=FALSE}
range(vec)
```
* Calculate the basal area of a tree of diameter 25 cm (the basal area is the area of a circle of the diameter of a tree), in m^2^

```{r, eval=FALSE}
pi * (25/200)^2
```


# Packages

Install the following packages from CRAN

**TO DO see the ones we will need for the data next week...**



