---
title: "Description of the wdData"
format:
  pdf
---

This is the description of the data wdData, that were available in a previous version of the package BIOMASS (v 2.1.11).
You can access these data in a csv file [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/5_graph/data/wdData.csv){preview-link="false"}.

# The global wood density database

## Description
The global wood density database (Chave et al. 2009, Zanne et al. 2009).

## Usage

```{r, eval = FALSE}
data("wdData")
```

## Format
A data frame with 16467 observations on the following 7 variables.

* family: a character vector indicating the family

* genus: a character vector indicating the genus

* species: a character vector indicating the species

* wd: a numeric vector of wood densities (g/cm^3)

* region: a character vector of regions (see getWoodDensity())

* referenceNumber: a numeric vector of reference numbers (bibliography)

* regionId: a character vector of region ids

## Details

This dataset is used in the function getWoodDensity(), to estimate a taxon-average wood density value.

## Source

Zanne et al. *Global wood density database*. Dryad. Identifier: http://datadryad.org/handle/10255/dryad.235 (2009).

## References
Chave et al. (2009) *Towards a worldwide wood economics spectrum*. Ecology letters 12:4, 351-366.

## Examples

```{r, eval = FALSE}
data(wdData)
str(wdData)
```

