---
title: "Exercices on session *Data visualisation*"
format:
  pdf
execute:
  echo: false 
  eval: false 
---

```{r, eval = TRUE}
#| message: false
library(tidyverse)
library(patchwork)
```

  
# Graphical exploration on wood density data

For this part, there is no need to improve the appearance of the plots.

Let's work with the *wdData* from a previous version of the package *BIOMASS* (v 2.1.11). These data are available [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/5_graph/data/wdData.csv){preview-link="false"}, and the metadata [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/5_graph/data/wdData.csv){preview-link="false"}.

* Read these data as a tibble

```{r}
wdData <- read_csv("data/wdData.csv")
```

* Make a histogram of wood density for each of the following three regions, using facets: South_America_(tropical), Australia/PNG_(tropical), Africa_(tropical)

```{r}
wdData %>% 
  filter(region %in% c("South_America_(tropical)", 
                       "Australia/PNG_(tropical)", 
                       "Africa_(tropical)")) %>% 
  ggplot(aes(x = wd)) + 
  geom_histogram() + 
  facet_wrap(~ region)
```

* Represent the same data but with a single density plot

```{r}
wdData %>% 
  filter(region %in% c("South_America_(tropical)", 
                       "Australia/PNG_(tropical)", 
                       "Africa_(tropical)")) %>% 
  ggplot(aes(x = wd, color = region)) + 
  geom_density()
```

* Compare the distribution of wood density for the 10 more common family on a single graph

```{r}
# do a boxplot because a density plot would be too much

abd_10 <- wdData %>% 
  count(family) %>% 
  arrange(desc(n)) %>% 
  slice_head(n = 10) 

wdData %>% 
  filter(family %in% abd_10$family) %>% 
  ggplot(aes(x = family, y = wd)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
* Show the number of observation per *family* and *regionId*

```{r}
wdData %>% 
  filter(family %in% abd_10$family) %>% 
  count(family, regionId) %>% 
  ggplot(aes(x = family, y = regionId)) + 
  geom_tile(aes(fill = n)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_c()
```
```{r}
wdData %>% 
  filter(family %in% abd_10$family) %>% 
  ggplot(aes(x = family, y = regionId)) + 
  geom_count()
```


# Iris data

Let's now work with the *iris* dataset (available from datasets, which is a package automatically loaded with R, your don't need to install it).

*"This famous (Fisher's or Anderson's) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica."*

```{r}
data(iris)
```

Reproduce the following graph made with the *iris* dataset.

The colours are *steelblue3*, *plum* and *darkolivegreen3* and the theme *bw*.

```{r, eval = TRUE}
#| fig-width: 8
#| fig-height: 8
#| message: false
p1 <- ggplot(iris, aes(x = Petal.Length, y=Petal.Width, 
                       color = Species)) + 
  geom_point() + 
  geom_smooth(method = lm) + 
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)",
       color = "",
       title = "Relationship between petal lenght and width") + 
  scale_x_continuous(breaks = 1:8) + 
  scale_y_continuous(breaks = seq(0, 2.5, by=0.5)) + 
  scale_color_manual(values = c("setosa" = "steelblue3",
                                "versicolor" = "plum",
                                "virginica" = "darkolivegreen3")) + 
  theme_bw() +
  theme(legend.position = "bottom")

p2 <- ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot() + 
  labs(x = "",
       y = "Petal length (cm)",
       color = "",
       title = "Distribution of petal length") + 
  scale_y_continuous(breaks = 1:8) + 
  scale_fill_manual(values = c("setosa" = "steelblue3",
                                "versicolor" = "plum",
                                "virginica" = "darkolivegreen3")) + 
  theme_bw() + 
  theme(legend.position = "none")

p3 <- ggplot(iris, aes(x = Species, y = Petal.Width, fill = Species)) +
  geom_boxplot() + 
  labs(x = "",
       y = "Petal width (cm)",
       color = "",
       title = "Distribution of petal width") + 
  scale_y_continuous(breaks = seq(0, 2.5, by=0.5)) +
  scale_fill_manual(values = c("setosa" = "steelblue3",
                                "versicolor" = "plum",
                                "virginica" = "darkolivegreen3")) + 
  theme_bw() + 
  theme(legend.position = "none")

p1 /(p2 | p3)
```


# For next time

* Make sure you remember what we saw on descriptive statistics in the session *Working with data*

* Install the following packages from CRAN: *multcompView*, *GGally*

* Make sure you have the following packages: *questionr*, *tidyverse*, *BIOMASS*
  
* Download the *RData* file available [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/6_basic_statistics/data/data_statistics_1.RData){preview-link="false"}
and store them in your raw data folder.