---
title: "Exercices on session *Working with data*"
format:
  pdf
execute:
  echo: false 
  eval: false 
---

# Working with data from the *Inventário Florestal Nacional*

-   Go to the website of the IFN ([here](https://snif.florestal.gov.br/pt-br/inventario-florestal-nacional-ifn/ifn-dados-abertos){preview-link="false"}) and download the data of the Rondônia for trees with a DBH≥10cm (don't forget to look at the metadata!)



- Import these data into R (make sure you look at the data structure before)

```{r}
data_Ron <- read.csv(file = "data/IFN-RO_arvores_dap_maior_10cm_20-04-2022.csv",
                     sep = ";",
                     dec = ",")
```

- How many variables are there? how many observations?

```{r}
ncol(data_Ron)
nrow(data_Ron)
dim(data_Ron)
```

- Look at the structure, everything is fine?

```{r}
str(data_Ron)
```

- Transform the following variable into factors: *Unidade amostral*, *estado de sanidade do fuste*, *qualidade do fuste*, 
*posição sociológica*, *hábito da planta* and check the structure again

```{r}
data_Ron$UA <- as.factor(data_Ron$UA) # Unidade amostral
data_Ron$SA <- as.factor(data_Ron$SA) # estado de sanidade do fuste
data_Ron$QF <- as.factor(data_Ron$QF) # qualidade do fuste
data_Ron$PS <- as.factor(data_Ron$PS) # posição sociológica
data_Ron$HAB <- as.factor(data_Ron$HAB) # hábito da planta
```

- Transform the following variable into boleans: *árvore fora da floresta*, *presença ou ausência de lianas*, 
*indivíduo com ou sem coleta botânica* and check the structure again

```{r}
data_Ron$AFF <- as.logical(data_Ron$AFF) # árvore fora da floresta
data_Ron$PL <- as.logical(data_Ron$PL) # presença ou ausência de lianas
data_Ron$MB <- as.logical(data_Ron$MB) # indivíduo com ou sem coleta botânica
```

- How many *Unidade amostral* is there?

```{r}
str(data_Ron$UA)

# alternative not seen in class:
length(unique(data_Ron$UA))
```


- Add a colum with the DBH in meter

```{r}
data_Ron$DAP_m <- data_Ron$DAP / 100
```

- What are the minimum, maximum and median height (*altura total*)

```{r}
min(data_Ron$HT)
max(data_Ron$HT)
# or 
range(data_Ron$HT)

median(data_Ron$HT)

# or 
summary(data_Ron$HT)
```

- Represent the distribution of DBH (in cm) using an histogram (*Bonus question for the foresters and forest ecologists: does this distribution make sense to you?*)

```{r}
hist(data_Ron$DAP)
```

- How many trees are out of the forest?

```{r}
table(data_Ron$AFF)
```
- Which percentage of trees have liana?

```{r}
questionr::freq(data_Ron$PL)
```

- Make a graphical representation (barplot or Cleveland plot) of the occurrences of health levels (*estado de sanidade do fuste*)

```{r}
tab <- table(data_Ron$SA)
tab <- sort(tab) # order in ascending order
barplot(tab)
```

```{r}
tab <- table(data_Ron$SA)
tab <- sort(tab) # order in ascending order
dotchart(as.vector(tab), # convert the table to a vector
         labels = names(tab))
```


# Find the problems

We are working with a fake data set of vegetation survey in the DF, that contain the following variables:

* *site*: name of the site

* *plot*: plot ID (unique identifier)

* *height_max*: height of the tallest tree on the plot (in m)

* *DBH_max*: DBH (diameter at breast height) of the largest tree on the plot (in cm)


```{r}
dt_pb <- data.frame(site = c("Brasilia", "Brasilia", "brasilia", "Taguatinga", "Taguatinga"), # 
                    plot = c(1:4, 4), #
                    height_max = c("10.3", "8.5", "9.6", "4,7", "9.8"), # 
                    DBH_max = c(20, 15, 31, 888, 12)) # 
write.csv(dt_pb,
          file = "data/dt_problem.csv",
          row.names = FALSE)
```

You can find this data set [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/3_manip_data/data/dt_problem.csv){preview-link="false"}.

There are 4 problems in this data frame, what are they?

```{r}
str(dt_pb)
summary(dt_pb)
```
<!-- There is a uncorrect decimal separator in height_max-->

```{r}
mode(dt_pb$height_max)
dt_pb$height_max
```
<!-- There is a problem of encoding in site-->

```{r}
dt_pb$site

# or better:
dt_pb$site <- as.factor(dt_pb$site)
levels(dt_pb$site)
# or 
table(dt_pb$site)
```


<!-- There is an incoherent value in DBH_max -->


```{r}
summary(dt_pb$DBH_max)

# or
range(dt_pb$DBH_max)
```
<!-- There is a duplicated plot ID -->

```{r}
table(dt_pb$plot)

# alternative not seen in the class
length(unique(dt_pb$plot)) == dim(dt_pb)[1]
```

# For next time

* Install the following package from CRAN: *tidyverse*

* Download the following data sets
[here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/4_tidyverse/data/func_traits.csv){preview-link="false"} and here [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/4_tidyverse/data/commercial_sp.csv){preview-link="false"} and store them in your raw data folder.

* Make sure you remember what we saw on factors and logical operators the session [Getting started with R](https://geraldinederroire.github.io/Course_R_Forest_Sciences/2_getting_started)