---
title: "Exercices on session *Working with data*"
format:
  pdf
execute:
  echo: false 
  eval: false 
---

# Working with data from the *Inventário Florestal Nacional*

-   Go to the website of the IFN ([here](https://dados.florestal.gov.br/pt_BR/dataset/ifn-dap-10-por-uf_disp-set2025){preview-link="false"}) and download the data of the Rondônia for trees with a DBH $\geq$ 10 cm (don't forget to look at the metadata!)



- Import these data into R (make sure you look at the data structure before)

```{r}
library(readxl)
data_Ron <- read_excel(path = "data/ifn_dap10_rondonia_disp-set2025.xlsx")
```

- How many variables are there? how many observations?

```{r}
ncol(data_Ron)
nrow(data_Ron)
dim(data_Ron)
```

- Look at the structure, everything is fine?

```{r}
str(data_Ron)
```

- Transform the following variable into factors: *Unidade de amostragem*, *Sanidade do indivíduo*, *Qualidade do fuste do indivíduo*, 
*Posição sociológica*, *Hábito dos indivíduos* and check the structure again

```{r}
data_Ron$UA <- as.factor(data_Ron$UA) # Unidade de amostragem
data_Ron$SA <- as.factor(data_Ron$SA) # Sanidade do indivíduo
data_Ron$QF <- as.factor(data_Ron$QF) # Qualidade do fuste do indivíduo
data_Ron$PS <- as.factor(data_Ron$PS) # Posição sociológica
data_Ron$HAB <- as.factor(data_Ron$HAB) # Hábito dos indivíduos
```

- Transform the following variable into boleans: *Arvore fora da floresta*, *Presença de Lianas*, 
*Coleta de material botânico* and check the structure again

```{r}
data_Ron$AFF <- as.logical(data_Ron$AFF) # Arvore fora da floresta
data_Ron$PL <- as.logical(data_Ron$PL) # Presença de Lianas
data_Ron$MB <- as.logical(data_Ron$MB) # Coleta de material botânico
str(data_Ron)
```

- How many *Unidade amostral* is there?

```{r}
length(levels(data_Ron$UA))

# alternative (but doesn't allow storing the result)
str(data_Ron$UA)

# alternative not seen in class:
length(unique(data_Ron$UA))
```


- Add a colum with the DBH in meter

```{r}
data_Ron$DAP_m <- data_Ron$DAP / 100
```

- What are the minimum, maximum and median height (*Altura total*)

```{r}
min(data_Ron$HT)
max(data_Ron$HT)
# or 
range(data_Ron$HT)

median(data_Ron$HT)

# or 
summary(data_Ron$HT)
```

- Represent the distribution of DBH (in cm) using an histogram (*Bonus question for the foresters and forest ecologists: does this distribution make sense to you?*)

```{r}
hist(data_Ron$DAP)
```

- How many trees are dead standing (*Sanidade do indivíduo* indicates *morta em pé*)?

```{r}
table(data_Ron$SA)

# alternative not seen in class
sum(data_Ron$SA == 4)
# or 
dim(data_Ron[data_Ron$SA == 4,])[1]
```

- Which percentage of trees are dead standing?

```{r}
questionr::freq(data_Ron$SA)
```

- Make a graphical representation (barplot or Cleveland plot) of the occurrences of health levels (*Sanidade do indivíduo*)

```{r}
tab <- table(data_Ron$SA)
tab <- sort(tab) # order in ascending order
barplot(tab)
```

```{r}
tab <- table(data_Ron$SA)
tab <- sort(tab) # order in ascending order
dotchart(as.vector(tab), # convert the table to a vector
         labels = names(tab))
```


# Find the problems

We are working with a fake data set of vegetation survey in the DF, that contain the following variables:

* *site*: name of the site

* *plot*: plot ID (unique identifier)

* *height_max*: height of the tallest tree on the plot (in m)

* *DBH_max*: DBH (diameter at breast height) of the largest tree on the plot (in cm)


```{r, echo=FALSE}
dt_pb <- data.frame(site = c("Brasilia", "Brasilia", "brasilia", "Taguatinga", "Taguatinga"), # 
                    plot = c(1:4, 4), #
                    height_max = c("10.3", "8.5", "9.6", "4,7", "9.8"), # 
                    DBH_max = c(20, 15, 31, 888, 12)) # 
write.csv(dt_pb,
          file = "data/dt_problem.csv",
          row.names = FALSE)
```

You can find this data set [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/3_manip_data/data/dt_problem.csv){preview-link="false"}.

There are 4 problems in this data frame, what are they?

```{r}
str(dt_pb)
summary(dt_pb)
```

```{r}
# There is a uncorrect decimal separator in height_max
mode(dt_pb$height_max)
dt_pb$height_max
```


```{r}
# There is a problem of encoding in site
dt_pb$site

# or better:
dt_pb$site <- as.factor(dt_pb$site)
levels(dt_pb$site)
# or 
table(dt_pb$site)
```

```{r}
# There is an incoherent value in DBH_max
summary(dt_pb$DBH_max)

# or
range(dt_pb$DBH_max)
```


```{r}
# There is a duplicated plot ID
table(dt_pb$plot)

# alternative
length(unique(table(dt_pb$plot))) ==  1

# alternative not seen in the class
length(unique(dt_pb$plot)) == dim(dt_pb)[1]
```


<!--
# For next time

* Install the following package from CRAN: *tidyverse*

* Download the following data sets
[here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/4_tidyverse/data/func_traits.csv){preview-link="false"} and [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/4_tidyverse/data/commercial_sp.csv){preview-link="false"} and store them in your raw data folder.

* Make sure you remember what we saw on factors and logical operators in the session [Getting started with R](https://geraldinederroire.github.io/Course_R_Forest_Sciences/2_getting_started)

-->
