---
title: "Statistics with R - slides not used"
author: Géraldine Derroire
institute: Cirad - UnB
date: last-modified
format: 
  revealjs:
    theme: solarized
    output-location: fragment 
    slide-number: true
    preview-links: true
    chalkboard: true
    link-external-icon: true
    link-external-newwindow: true
    incremental: true
execute:
  echo: true   
  warning: true
  message: true 
  cache: true
editor: 
  markdown: 
    wrap: sentence
---

```{r}
library(ggplot2)
```


## Log-normal distribution {.smaller}

$x \sim log\mathcal{N}(\mu, \sigma^2)$ if $log(x) \sim \mathcal{N}(\mu, \sigma^2)$

[We use the function [rnorm]{style="color:indianred;"} to randomly draw data from a normal distribution:]{style="font-size: 30px"}

::::: columns
::: {.column width="60%"}


```{r}
#| warning: false
dt1 <- data.frame(x = rlnorm(n = 100000, mean = 0, sd = 1))
dt2 <- data.frame(x = rlnorm(n = 100000, mean = 1, sd = 1))

ggplot() + 
  geom_density(data = dt1, aes(x = x), 
               color = "blue") + 
  geom_density(data = dt2, aes(x = x), 
               color = "green") + 
  scale_x_continuous(limits =c(0,20)) +
  theme_minimal()
```
:::

::: {.column width="40%"}

* The values are stricty positive.

* The variance increases when the mean increases. 

:::
:::::


## Bernoulli and Binomial distribution {.scrollable}

$$\mathcal{B}(n, p)$$

[The binomial distribution represent the number of successes when performing ***n*** independent experiment having a constant probability of success ***p***.]{style="font-size: 30px"}

[It is a discrete distribution taking only **positive integer values**.{style="font-size: 30px"}

> [[It can be use for proportion: *eg* proportion of trees browsed in a plot.]{style="font-size: 30px"}]{.fragment}

::: fragment
[The Bernoulli distribution is a special case of the binomial distribution where ***n=1***.]{style="font-size: 30px"}

It is a discrete distribution taking only the values **0 (failure) or 1 (success)**.]{style="font-size: 30px"}
:::

> [It can be use for presence/abscence and success/failure data: *eg* probability that a tree is browsed.]{style="font-size: 30px"}]{.fragment}



## Bernoulli and binomial distribution {.scrollable}

We use the function [rbinom]{style="color:indianred;"} to randomly draw data from a binomial or a Bernoulli distribution:

::::: columns
::: {.column width="33%"}
```{r}
dt1 <- data.frame(x = 
          rbinom(n = 100000, 
                 size  = 20, 
                 p = 0.5))
ggplot() + 
  geom_bar(data = dt1, 
           aes(x = x)) +
  theme_minimal()
```
:::

::: {.column width="33%"}
```{r}
dt2 <- data.frame(x = 
          rbinom(n = 100000, 
                 size = 20, 
                 p = 0.7))
ggplot() + 
  geom_bar(data = dt2, 
           aes(x = x)) +
  theme_minimal()
```
:::

::: {.column width="33%"}
```{r}
# Bernoulli distribution
dt3 <- data.frame(x = 
          rbinom(n = 100000, 
                 size = 1, 
                 p = 0.5))
ggplot() +
  geom_bar(data = dt3, 
           aes(x = x)) +
  theme_minimal()
```
:::
:::::


## Poisson

TO DO OR TO MOVE TO SESSION 2

discrete

cf Eric: 1 slide et Marchand

## To go further on distribution

TO PUT IN THE RESOURCES OF SESSION 2/3

distribution zoo + wikipedia pages (in english)

