---
title: "Exercices on session *Statistics with R session 1/3*"
format:
  pdf
execute:
  echo: false 
  eval: false 
---

Load the package tidyverse

```{r}
library(tidyverse)
```

# Admission in graduate school in UC Berkeley

*We want to know if the admissions in graduate school in the Department A of UC Berkeley depends on the gender.*


* The data are called *UCBAdmissions* and are available in base R. Load the data and look at their structure

```{r}
data(UCBAdmissions)
str(UCBAdmissions)
```

* Get the data for department A (the first element of the array so you need to use the following command)

```{r, echo=TRUE}
UCBAdmissions[, , 1]
```

```{r}
dt_deptA <- UCBAdmissions[, , 1]
```

* These data give the number of admitted or rejected people per gender. How is this type of table called?

<!-- a contingency table -->

* Perform a $\chi^2$ test to see if the admission depends on the gender, and interpret it

```{r}
res_chi2 <- chisq.test(dt_deptA)
res_chi2
```

<!-- low p-value => yes the admission depends on the gender-->

* Check the expected frequency under the null hypothesis. Is the $\chi^2$ test reliable?

```{r}
res_chi2$expected
```

<!-- yes reliable (all expected > 5)-->

* Get the residuals and interpret them

```{r}
questionr::chisq.residuals(dt_deptA)
```

<!-- over-repressentation of the female in the admitted => female more admitted than male-->

* Represent the residuals graphically

```{r}
mosaicplot(dt_deptA, 
           shade = TRUE,
           main = "") # remove main title
```


# Weight of cabbages from different cultivars

*We want to know if the weight of cabbages planted on a certain day depends on the cultivar.*

* Load the dataset called *cabbages* in the package *MASS*

```{r}
library(MASS)
data("cabbages")
str(cabbages)
```

* Make a dataset containing only the cabbages planted at date *d21*

```{r}
dt_cab <- cabbages %>% filter(Date == "d21")
```


* Make a boxplot to graphically explore if the weight depends on the cultivar. What do you think? 

```{r}
dt_cab %>% 
  ggplot(aes(x = Cult, y = HeadWt)) + 
  geom_boxplot()
```


* Check the normality of the distribution, using a test and graphically

```{r}
# make a dataset per cultivar
dt_c39 <- dt_cab %>% filter(Cult == "c39")
dt_c52 <- dt_cab %>% filter(Cult == "c52")
```

```{r}
# Shapiro test of normality (H0: the data are normally distributed)
shapiro.test(dt_c39$HeadWt)
shapiro.test(dt_c52$HeadWt)
# not significant so the data are normally distributed
```

```{r}
# Q-Q plots
qqnorm(dt_c39$HeadWt)
qqline(dt_c39$HeadWt)

# Q-Q plots
qqnorm(dt_c52$HeadWt)
qqline(dt_c52$HeadWt)

# very little points, so hard to interpret...
```


* Perform a t-test

```{r}
res_ttest <- t.test(dt_cab$HeadWt ~ dt_cab$Cult)
res_ttest
```

* Get the p-value, the mean estimated difference between the two groups and its confidence interval. What do you conclude?

```{r}
res_ttest$p.value
```

<!-- The difference is significant-->

```{r}
cab_est <- res_ttest$estimate
abs(cab_est[1] - cab_est[2])
# or 
abs(cab_est[2] - cab_est[1])
# The name of the vector of result is not correct
```

<!-- The difference between group is big (compared to the mean per group) -->

```{r}
res_ttest$conf.int
# 0.95 is the confidence level
```

<!-- big confidence interval -->

# Effect of the planting date on the content of vitamin C in cabbages

*We want to know if the content of vitamin C in cabbages of the cultivar c52 depends on the planting date.*

* Make a dataset containing only the cabbages of the cultivar *c52*

```{r}
dt_cab2 <- cabbages %>% filter(Cult == "c52")
```

* Make a boxplot to explore the question graphically. What do you think?

```{r}
dt_cab2 %>% 
  ggplot(aes(x = Date, y = VitC)) + 
  geom_boxplot()
```

* Perform an ANOVA

```{r}
aov_cab <- aov(VitC ~ Date, data = dt_cab2)
summary(aov_cab)
```

<!-- the test is significant => the content of vitamin C depend on the planting date -->

* Check the model assumptions

```{r}
# check homoscedasticity
plot(aov_cab, which = 1)
# => OK
```

```{r}
# check normality of the residuals
plot(aov_cab, which = 2)
# => OK
```

* Get the estimated mean of each groups 

```{r}
coef(aov_cab)

# d16
coef(aov_cab)[1]

# d20
coef(aov_cab)[1] + coef(aov_cab)[2]

# d21
coef(aov_cab)[1] + coef(aov_cab)[3]
```

* Perform a Tukey test see which groups are different, and interpret it

```{r}
tukey_cab <- TukeyHSD(aov_cab)
tukey_cab
```

<!-- d21 differs from the two others --> 

* Represent the results of the ANOVA graphically

```{r}
# get the letters
cab_let <- multcompView::multcompLetters4(aov_cab, tukey_cab)

# put them in a tibble
cab_let_tb <- tibble(
  Date = names(cab_let$Date$Letters),
  letter = as.character(cab_let$Date$Letters)
  )

# make a tibble with the dates, the letters and the 75th percentile
cab_let_pos <- dt_cab2 %>% 
  group_by(Date) %>% 
  summarise(q_75 = quantile(VitC, probs = 0.75, na.rm = TRUE)) %>% 
  inner_join(cab_let_tb)

ggplot() +
  geom_boxplot(data = dt_cab2, 
               aes(x = Date, y = VitC, fill = Date),
               show.legend = FALSE) +
  geom_text(data = cab_let_pos,
            aes(x = Date, y = q_75, # position of the label
                label = letter), # labels are the letters
            size = 5, # size of the labels
            vjust = -1, hjust = -1) + # adjusts position of the labels
  scale_fill_brewer(palette = "Blues") + 
  theme_bw()
```


# Brain and body weights in mamals

*We want to know if there is a correlation between the brain and the body weights in mamals.*

* Load the dataset called *mammals* in the package *MASS*

```{r}
data(mammals)
str(mammals)
```

* Get the Pearson correlation coefficient

```{r}
cor(mammals$body, mammals$brain)
```
<!-- very high correlation-->

* Do a Pearson correlation test and interpret it

```{r}
cor.test(mammals$body, mammals$brain,
         method = "pearson")
```
<!-- There is a strong significant positive relationship between body and brain mass-->


# For next time

* Install the following packages from CRAN: *ggeffects*, *gtsummary*, *ggstats*, *lindia*, *car*, *AICcmodavg*

* Make sure the following packages are already installed: *tidyverse*, *MASS* 

* Download the *RData* file available [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/6_basic_statistics/data/data_statistics_2.RData){preview-link="false"}
and store them in your raw data folder.