---
title: "Exercices on session *Tidyverse 1/2*"
format:
  pdf
execute:
  echo: false 
  eval: false 
---

```{r, message=FALSE}
library(BIOMASS)
library(tidyverse)
```

For all the following exercises, use functions from the tidyverse whenever possible, and try to make one pipe per question.

# Karnataka forest dataset

You are going to work on data from forests of the central Western Ghats of India.
These data are available in the package BIOMASS, and are called 
KarnatakaForest.

* Load the data (don't forget to look at the metadata in the help of the package). 
Rename the data to facilitate the writing of the script.

```{r}
data(KarnatakaForest)
dt_WG <- KarnatakaForest
rm(KarnatakaForest)
```

* Get an overview of the data and check their class.

```{r}
str(dt_WG)
summary(dt_WG)
class(dt_WG)
```

* Transform the data to a tibble. 

```{r}
dt_WG <- as_tibble(dt_WG)
str(dt_WG)
```

* Transform all non-numeric variables to a factor.

```{r}
dt_WG <- dt_WG %>% 
  mutate(plotId = as.factor(plotId),
         treeId = as.factor(treeId),
         family = as.factor(family),
         genus = as.factor(genus),
         species = as.factor(species))
```

# Diameters

* Rename the column *D* to *diam*

```{r}
dt_WG <- dt_WG %>% rename(diam = D)
```

* What is the smallest diameter of the whole dataset (make sure you extract the value, not only read it in a output)?

```{r}
diam_min <- dt_WG %>% 
  slice_min(diam) %>% # get only the rows with the smallest diameter
  select(diam) %>% # keep only the variable diameter
  distinct() # keep only one unique value
diam_min
```


* What are the full species names (genus and species) of the 5 biggest trees?

```{r}
big5 <- dt_WG %>% 
  arrange(desc(diam)) %>% # sort the tibble by decreasing order of diameter
  slice_head(n=5) %>% # take the first 5 rows
  select(genus, species) # select the columns 
big5
```

* How many different species are amongst these five biggest trees?

```{r}
sp_big5 <- big5 %>% 
  distinct() %>% # select unique rows
  count() # count them

# NB: this is different from
# big5 %>% count(genus, species)
# which counts the number of tree for each species
```

* Create a table (tibble) with the median diameter per plot.

```{r}
dt_med_diam <- dt_WG %>% 
  group_by(plotId) %>% # group by plotId
  summarize(median_d = median(diam)) # takje the median by plotId
dt_med_diam
```

* What is the most common family in the whole dataset?

```{r}
com_fam <- dt_WG %>%
  count(family) %>% # count the number of trees per family
  slice_max(n) # get the one with the highest number of tree
com_fam
```

* Create a tibble with the number of trees per family and per plot for plots BSP1 and BSP2, and order it by decreasing number of individuals per family and per plot. 

```{r}
dt_sub <- dt_WG %>% 
  filter(plotId %in% c("BSP1", "BSP2")) %>% # keep only the two plots of interest
  count(plotId, family) %>% # count the number of trees per family and plot
  arrange(plotId, desc(n)) # sort them
view(dt_sub)
```

* To how many families do trees that have a diameter in the fourth quartile (bigger than the 3rd quartile) of the whole dataset belong?

```{r}
n_75 <- dt_WG %>% 
  filter(diam > quantile(diam, probs = 0.75)) %>% # select trees with a diameter in the fourth quartile
  select(family) %>% # keep only the variable family
  distinct() %>% # keep only a unique row for each family
  count() # count the rows
n_75
```


# For next time

* Install the following package from CRAN: *janitor*

* Make sure you remember what we saw on factors and logical operators during the session [Getting started with R](https://geraldinederroire.github.io/Course_R_Forest_Sciences/2_getting_started)

* Download the *RData* file available [here](https://geraldinederroire.github.io/Course_R_Forest_Sciences/4_tidyverse/data/data_tidyverse_2.RData){preview-link="false"}
and store them in your raw data folder.

